---
title: "Introduction to svcnvplys"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to svcnvplys}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r style, echo = FALSE, results = 'asis'}
BiocStyle::markdown()
```
  
```{r include=FALSE}
require(taRifx)
require(data.table)
require(tidyr)
require(circlize)
require(GenomicRanges)
require(D3GB)
```

## svncvplus: R toolkit for the downstream analyses of structural variantions

`svncvplus` is an R package designed for integrative analyses of somatic DNA copy number variations (CNV) and other structural variants (SV).`svcnvplus` toolkit comprises multiple analyticall and visualization that can be applied to large datasets from cancer patients (e.g. [TCGA](https://www.cancer.gov/about-nci/organization/ccg/research/structural-genomics/tcga)) and cencer cell lines (e.g. [CCLE](https://portals.broadinstitute.org/ccle)). 
  
CNV data can be derived from genotyping and CGH arrays, as well as next generation sequencing; different segmentation algorithms are used to obtain dosage variations (duplications and deletions) across the genome. Alternatively SV calls can be inferred from discordantly aligned reads from whole genome sequencing (WGS) using different algorithms (e.g [manta](https://github.com/Illumina/manta)[], [lumpy](https://github.com/arq5x/lumpy-sv), etc).
  
SV calls provide linkage information from discordantly aligned reads and read pairs, allowing the discovery of chromosomal translocations and variants that do not necesarily involve dosage change, such as inversions and insertions. Segmentation CNVs and alignment based SV calls produce orthogonal as well as complementary results. The integration of both data types can by highly informative to understand the somatic alterations driving many cancers and is essential to characterize complex chromosomal alterations such as chromothripsis and chromoplexy.
  
Most currently available cancer genomics datasets incorporate CNV characterization whereas SV derived from WGS are more scarce. For this reason, `svcnvplus` tools implements functions that work with both data types separately.


## Index:

* [Install svcnvplus](#install-svcnvplus)
* [Input data](#input-data)
* [Validate data types](#validate-data-types)
    * [Validate segmentation data format](#validate-segmentation-data-format)
    * [Validate structural variant data format](#validate-structural-variant-data-format)
* [CNV analysys and visualization](#cnv-analysys-and-visualization)
    * [CNV frequency plot](#cnv-frequency-plot)
    * [Percent genome change](#percent-genome-change)
* [Breakpoint burden](#breakpoint-burden)
* [Co-localization of breakpoints](#co\-localization-of-breakpoints)
* [Identification of shattered regions](#identification-of-shattered-regions)
    * [Chromosome shattering using segmentation data only](#chromosome-shattering-using-segmentation-data-only)
    * [Chromosome shattering using segmentation and SV data](#chromosome-shattering-using-segmentation-and-sv-data)
* [Recurrently shattered regions](#recurrently-shattered-regions)
    * [Recurrently shattered regions plot](#recurrently-shattered-regions-plot)
* [Recurrently altered genes](#recurrently-altered-genes)
    * [Gene level CNV](#gene-level-cnv)
    * [Recurrently altered genes overlapping with CNV breakpoints](#recurrently-altered-genes-overlapping-with-cnv-breakpoints)
    * [Recurrently altered genes overlapping with structural variants](#recurrently-altered-genes-overlapping-with-sv-breakpoints)

------------

## Install svcnvplus

Install development version from GitHub
```{r message=FALSE, eval=FALSE}
devtools::install_github("gonzolgarcia/svcnvplus")
```

## Input data

Two data types are allowed:
  
__CNV segmentation data:__ 6 columns are required in the folowing order: `sample`, `chrom`, `start`, `end`, `probes` & `segmean`. Most algorithms studying CNVs produce segmented data indicating genomic boundaries and the segment mean copy number value (segmean); `svcnvplus` assumes CNV expresed as log-ratios: __e.g.:__ $\log2(tumor/normal)$ Those values do not necesarily represent entire copy number states as many samples may contain admixture or subclonal populations.
  
__Structural Variant calls:__ 8 columns are required in the folowing order: `sample`, `chrom1`, `pos1`, `strand1`, `chrom2`, `pos2`, `strand2` & `svclass`. SV calls are obtained from WGS by identifying reads and read-pairs that align discordantly to the reference genome. The types accepted in the svclass field are: duplication(DUP), deletion(DEL), inversion(INV), insertion(INS), translocation(TRA) and breakend(BND) for undefined variants.
  
All functions acceps multiple samples. Functions that make use of both VNC and SV calls expect a commn set of ids in the `sample` field.
  
In order to explore the  functionalities of svcnvplus, two data.frames have been included with the package:
  
* CCLE lung cancer derived cell lines; This data.frames contain information about CNV segments and structural variants respectively for lung cancer derived cell lines. This data was obtained from <https://depmap.org/portal/download/>.
    * `svcnvplus::segdat_lung_ccle`
    * `svcnvplus::svdat_lung_ccle`.  
* TARGET neuroblastoma dataset based on cComplete Genomics whole genome sequencing     
    * `svcnvplus::nbl_segdat`
    * `svcnvplus::nbl_svdat`.  
Both datasets are `lazy` loaded with `svcnvplus`

```{r message=FALSE}
library(svcnvplus)
head(nbl_segdat)
head(nbl_svdat)
```

## Validate data types

* `validate.seg()` check format for segmentation data.frame
* `validate.sv()` chrek format of a Structural variant data.frame

### Validate segmentation data format

Validate and format segmentation data.frame to be used by svcnvplus package toolkit

```{r}
segdf <- validate.seg(nbl_segdat)
head(segdf)
```


### Validate structural variant data format 

Validate and format structural variant data.frame to be used by svcnvplus package toolkit

```{r}
svdf <- validate.sv(nbl_svdat)
head(svdf)
```

## CNV analysys and visualization

### CNV frequency plot

Visualization of CNV gain/loss frequencies across the genome; aggregates samples for a given genomic window size, which copy number log-ratio differ from 0. The thresshold `fc.pct` is represented as percentage (e.g. 0.2 -> 20% fold change with the referece).

```{r plot1,  fig.width=9, fig.height=4, fig.align='center', fig.cap = "Genome wide CNV frequencies", message=FALSE}
cnv_freq <- cnv.freq.plot(segdf, fc.pct = 0.2)  # plot cnv frequencies
```

```{r message=FALSE}
head(cnv_freq$freqsum)  # data.frame contains every genomic bin 

```


### Percent genome change

Per sample measure of genome instability; calculates what percentage the genome’s copy number log2-ratio differs from 0 iven a `fc.pct` thershold.

```{r message=FALSE}
pct_change <- pct.genome.changed(segdf, fc.pct = 0.2)
```

### Chromosome arm CNV

The function `chr.arm.cnv` obtaines the segment weighted average log-ratio for each chromosome arm returned in a matrix format.

```{r plot1.2, fig.width=9, fig.height=7, fig.align='center', fig.cap = "SV versus CNV breakpoint burden",  message=FALSE}
charm.mat <- chr.arm.cnv(segdf, genome.v = "hg19", verbose = FALSE)
require(gplots2,quietly = TRUE,warn.conflicts = FALSE)

heatmap.2(charm.mat[,],Rowv=NA,trace='none',cexCol=.5, lhei=c(0.25,1), dendrogram='col', key.title="Copy number",
        col=colorRampPalette(c("blue","white","red"))(256))
```


## Breakpoint burden

In addition to percentage of genome changed, we can measure the total burden of breakpoints derived from CNV segments and SV calls. 

```{r plot2, fig.width=9, fig.height=5, fig.align='center', fig.cap = "SV versus CNV breakpoint burden",  message=FALSE}
# define breakpoints from SV data
sv_breaks  <- sv.breaks(svdf)  

# define breakpoints from seg data based on certain CNV change cutoff
cnv_breaks  <- seg.breaks(segdf,fc.pct = 0.2,verbose=FALSE)  

# scatter plot comparing CNV and SV breakpoint burden for a set of common samples
common_samples <- intersect(names(sv_breaks$brk.burden),names(cnv_breaks$brk.burden))

dat1 <- log2(1+cbind(sv_breaks$brk.burden[common_samples],
                    cnv_breaks$brk.burden[common_samples]))

dat2 <- log2(1+cbind(pct_change, 
                     cnv_breaks$brk.burden[names(pct_change)]))

par(mfrow=c(1,2))
plot(dat1, xlab="SV burden", ylab="CNV breakpoint burden")
legend("topright",paste("cor=",cor(dat1)[1,2], sep=""))
plot(dat2, xlab="percentage genome changed", ylab="CNV breakpoint burden")
legend("topright",paste("cor=",cor(dat2)[1,2], sep=""))

```

## Co-localization of breakpoints

Both CNV segmentation profiles and SV calls produce orthogonal results for variants that involve dosage changes (duplications and deletions). The function `match.breaks` compare both approaches by identifying colocalizing breakpoints. This function can also be used to compare two sets of CNV brekpoints obtaind from different algorithms or SV callers since the format of both CNV and SV breaks objects have the same format.  

```{r plot3,  fig.width=9, fig.height=5, fig.align='center', fig.cap = "Common breakpoints by sample",  message=FALSE}

common.breaks <- match.breaks(cnv_breaks, sv_breaks, 
                              maxgap=50000, verbose=FALSE)

# average percentage of colocalizing breaks
restab <- data.frame(common.breaks$restab)[order(common.breaks$restab$total.brk2),]
m2 <- sprintf("%.1f",100*mean(restab$matched.brk2/restab$total.brk2)) 

# Plot the proportion of SV breakpoints that have colocalizing CNV breakpoints
barplot(rbind(restab$matched.brk2, restab$total.brk2 - restab$matched.brk2),
        border=NA,las=2,xlab="",horiz=FALSE,cex.main=.7,cex.names=.4, names=rownames(restab))
legend("top",paste("SV breaks matched by CNV breaks\n","Average = ",m2,"%",sep=""),bty='n')
grid(ny=NULL,nx=NA)
```


## Identification of shattered regions

Complex chromosomal rearrangements such as chromothripsis and chromoplexy are widespread events in many cancers and may have important pathogenic roles. `svcnvplus` incorporates tools to map and visualize shattered regions across multiple samples.

We used LUNG cancer cell line profiles from the CCLE in order to illustrate these tools:

Validate segmentation and SV data.frames
```{r message=FALSE}
segdf <- validate.seg(segdat_lung_ccle)
svdf <- validate.sv(svdat_lung_ccle)
```

### Chromosome shattering using CNV segmentation data only

The whole genome is binned into user defined `window.size` (Mb) and slided by `slide.size` (Mb) in order to identify regions with high CNV breakpoint density. Two cutoffs are considered for each genomic bin:
  
* num.breaks = the minimum number of breakpoints
* num.sd = the number of standard deviations above the average within a sample
  
In addition, we evaluate the interquantile average of the distance between breakpoints in a given region:
  
* dist.iqm.cut (default = 150000 b)

```{r message=FALSE}
shatt_lung_cnv <- shattered.regions.cnv(segdf, fc.pct = 0.2, clean.brk = 4, window.size = 10,
                                        min.num.probes = 3, slide.size = 2,num.breaks = 8, 
                                        num.sd = 5, dist.iqm.cut = 150000,verbose=FALSE)
shatt_lung_cnv$regions.summary$A549_LUNG
```


### Chromosome shattering using segmentation and SV data

Analogously we can combine CNVs and SVs breakpoins to obtain a more robust evaluation of chromosome shattering (see `?shattered.regions`) 
In addition SVs provide linkage for each SV breakpoint pair which allow for an additional parametter:
  
* interleaved.cut the minimun percentage (0-1) of interleaved SVs
    
```{r message=FALSE}
shatt_lung <- shattered.regions(segdf, svdf, fc.pct = 0.05,  min.num.probes = 3, clean.brk = 4,
                                window.size = 10, slide.size = 2, num.seg.breaks = 6, 
                                num.seg.sd = 5, num.sv.breaks = 6, num.sv.sd = 5, 
                                num.common.breaks = 2, num.common.sd = 0, interleaved.cut = 0.33,
                                dist.iqm.cut = 100000,verbose=FALSE)
shatt_lung$regions.summary$NCIH522_LUNG
```

Circos plotting is available via [circlize](https://cran.r-project.org/web/packages/circlize/index.html) package wrapper function `circ.chromo.plot`:

```{r plot4, fig.width=9, fig.height=5, fig.align='center', fig.cap = "Circos plot representing c LUNG cancer cell lines with chromothripsis",  message=FALSE}
par(mfrow=c(1,2))
circ.chromo.plot(shatt_lung,sample.id = "SCLC21H_LUNG")
circ.chromo.plot(shatt_lung,sample.id = "NCIH510_LUNG")
```


### Recurrently shattered regions

We evaluate the null hipothesis that shattered regions occour throughout the genome at random; To this end we first create an empirical null distribution based on the sample set under study. The null is then compared with the observed distribution to obtain empirical p-values corrected for multiple hypothesis testing. The corrected p-calues under statistical significance define regions  under selection presure for chromosome shattering.

```{r message=FALSE}
fdr.test <- freq.p.test(shatt_lung_cnv$high.density.regions.hc, method="fdr", p.cut = 0.01)
```

### Recurrently shattered regions plot

We can visualize the aggregate map of shattered regions for all samples with `shattered.map.plot`. 

```{r plot5,  fig.width=9, fig.height=4, fig.align='center', fig.cap = "Recurrently shattered regions map", message=FALSE}
shattered.map.plot(shatt_lung_cnv, fdr = fdr.test$freq.cut)
```

And finally collect groups of samples with shattered chromosomes in recurrent regions as defined by a given fdr cutoff.

```{r message=FALSE}
# obtain genomic bins within above the FDR cutoff
freq.matrix <- apply(shatt_lung_cnv$high.density.regions.hc,2,sum)
textRegions <- names(which(freq.matrix >= fdr.test$freq.cut))
hitRegions <- remove.factors((data.frame(do.call(rbind,strsplit(textRegions," ")))))
hitRegions[,2] <- as.numeric(hitRegions[,2])
hitRegions[,3] <- as.numeric(hitRegions[,3])
colnames(hitRegions) <- c("chr","start","end")
rownames(hitRegions) <-textRegions

# collapes contiguous bins into unique regions
bins2remove <- c()
for(i in 2:nrow(hitRegions)){ 
  if(hitRegions[i,"chr"] == hitRegions[i-1,"chr"] ){
    if(hitRegions[i,"start"] < (hitRegions[i-1,"end"])){
      hitRegions[i,"start"] <- hitRegions[i-1,"start"]
      bins2remove <- c(bins2remove,textRegions[i-1])
    }
  }
}
hitRegionsPost<- hitRegions[setdiff(rownames(hitRegions),bins2remove),]

require(GenomicRanges)
hitRegions_gr <- with(hitRegions, GRanges(chr, IRanges(start=start, end=end)))
hitRegionsPost_gr <- with(hitRegionsPost, GRanges(chr, IRanges(start=start, end=end)))
hits <-GenomicAlignments::findOverlaps(hitRegionsPost_gr,hitRegions_gr)

regList <- list()
for(i in unique(queryHits(hits))) regList[[paste(hitRegionsPost[i,],collapse=" ") ]] <- textRegions[subjectHits(hits)[which(queryHits(hits) == i)]]

# obtain the genomic bins with maximum number of samples
regListPeak <- lapply(regList, function(x) 
    names(which(freq.matrix[x] == max(freq.matrix[x]))))
# collect samples with shattered region in the peaks 
regListPeakSamples <- lapply(regListPeak, function(x) 
    names(which(apply(cbind(shatt_lung_cnv$high.density.regions.hc[,x]),1,sum) > 0)))
```

Beyond this point the user can test case/control hipothesys for chromosome shattering of specific genomic regions within the dataset under study.


## Recurrently altered genes

Somatic pathogenic variants are characterize by presenting in recurrent patterns . Evaluating the recurrence of structural variations involve challenges as their interpretation more complicated than other variant types (e.g. SNVs). We evaluate the recurrence of structural variants using two criteria: dosage changes at the gene level and analysis of breakpoints overlapping with known genes.

### Gene level CNV

Generates a matrix with gene level CNVs from a segmentation file and obtain the a ranking of amplifications and deep deletions (cutoff = 2 => * copies; -2 => 0.5 copies).

```{r message=FALSE}
segdf_clean <- clean.cnv.artifact(segdf, verbose=FALSE,n.reps = 4)  # remove likely artifacts from segmentation data
gene_cnv <- gene.cnv(segdf_clean,genome.v = "hg19",fill.gaps = TRUE,verbose=FALSE)
amplified <- apply(gene_cnv$cnvmat, 1, function(x) which(x > 2))
deepdel <- apply(gene_cnv$cnvmat, 1, function(x) which(x < -2))
```

```{r plot6,  fig.width=9, fig.height=5, fig.align='center', fig.cap = "Recurrently altered genes with overlapping CNV breakpoints", message=FALSE}
par(mfrow=c(1,2),mar=c(4,7,1,1))
barplot(sort(unlist(lapply(amplified,length)),decreasing=TRUE)[20:1],col="red",
        las=1,main="Amplified genes",horiz=TRUE,xlab="#samples")
barplot(sort(unlist(lapply(deepdel,length)),decreasing=TRUE)[20:1],col="blue",
        las=1,main="Candidate homozigously deleted genes",horiz=TRUE,xlab="#samples")
```

### Recurrently altered genes overlapping with CNV breakpoints

Instead of focusing on high-level dosage changes, we evaluate whether CNV breakpoints overlap with known genes or upstream regions (gene level CNVs are studied [above](#gene-level-cnv)). 'We input a `cnv.break.annot` evaluates segmentation data and produces

```{r plot7,  fig.width=9, fig.height=5, fig.align='center', fig.cap = "Recurrently altered genes with overlapping CNV breakpoints", message=FALSE}
results_cnv <- cnv.break.annot(segdf_clean, fc.pct = 0.2, genome.v="hg19",clean.brk = 8,upstr = 50000,verbose=FALSE)
par(mfrow=c(1,2),mar=c(4,7,1,1))
barplot(sort(unlist(lapply(results_cnv$disruptSamples,length)),decreasing=TRUE)[20:1],
        las=1,main="Gene coding region disrupted",horiz=TRUE,xlab="#samples")
barplot(sort(unlist(lapply(results_cnv$upstreamSamples,length)),decreasing=TRUE)[20:1],
        las=1,main="Gene upstream region disrupted",horiz=TRUE,xlab="#samples")
```

### Recurrently altered genes overlapping with structural variants calls

SV calls do not incorporate dosage information, therefore we study the localization of breakpoints with respect to known genes. The annotation identifies small segmental variants overlaping with genes. For translocations (TRA) and large segmenral variants (default > 200Kb) only the breakpoint overlap with genes are considered.
The output returns a list of genes and associated variants that can be retrieved for further analyses. In addition every gene is associated via list to the sample ids harboring variants

```{r plot8,  fig.width=9, fig.height=5, fig.align='center', fig.cap = "Recurrently altered genes with overlapping SV breakpoints", message=FALSE}
results_sv <- sv.break.annot(svdf, sv.seg.size = 200000, genome.v="hg19",upstr = 50000, verbose=FALSE)
par(mfrow=c(1,2),mar=c(4,7,1,1))
barplot(sort(unlist(lapply(results_sv$disruptSamples,length)),decreasing=TRUE)[20:1],
        las=1,main="Coding region disrupted",horiz=TRUE,xlab="#samples")
barplot(sort(unlist(lapply(results_sv$upstreamSamples,length)),decreasing=TRUE)[20:1],
        las=1,main="Upstream region disrupted",horiz=TRUE,xlab="#samples")
```


